Isomorphic Interpreters from Logically Reversible Abstract Machines

### 2012-09-15 iso-int/rc-iso-int.tex
several corrections

\paragraph*{Universality.} 
As an example, consider the implementation of the Toffoli gate below, which
establishes that {{langRevT}} is complete for combinational circuits. This
encoding uses the type {{1+1}} as {{bool}} and values {{left ()}} and
{{right()}} as {{true}} and {{false}}:
\begin{center}
  \includegraphics{diagrams/toffoli.pdf}
\end{center}
\noindent Previous work~\cite[Sec.~5]{James:2012:IE:2103656.2103667}
establishes that {{langRevT}} is Turing complete by showing the compilation
of a Turing complete language --- a first-order language with numbers and
iteration --- to {{langRevT}}.

@@

\vspace{-20pt}
\begin{small}
\begin{multicols}{2}  
%subcode{bnf} include main
% Numbers, n, m,p = 0 | n + 1
% Machine states = <e, e, e>

@@

\paragraph*{Stuck States.}
The type systems of most languages are not expressive enough to encode the
precise domain and range of a function. For example, in most typed languages,
division by zero is considered type-correct and the runtime system is
required to deal with such an error. A stuck state of an abstract machine is
just a manifestation of this general problem. The common solutions are:
\begin{itemize}
\item \emph{Use a more expressive type system.}  One could augment
  {{langRevT}} with a richer type system that distinguishes non-zero numbers
  from those that can be zero, thereby eliminating the {{sub1 0}} situation
  entirely. Similarly, in the meta-circular interpreter in
  Sec.~\ref{sec:langrevt-int}, one could use generalized abstract data types
  (GADTs)~\cite{phantom,Xi:2003:GRD:604131.604150}) to eliminate the stuck
  states.

@@

We show the generality of our construction by applying it to two non-trivial
examples: a tree traversal and a meta-circular interpreter for
{{langRevT}}.

@@

Deriving {{langRevT}} combinators from abstract machines can sometimes
be used as an efficient indirect way of programming in {{langRevT}}.
Say we wished to build a {{langRevT}} combinator to map {{nat}} to a
parity encoded version {{par = @@x.1+(1+x)}}. Since {{par}} is not a
fixed unfolding of {{nat}}, it is not immediately apparent how such a
combinator can be constructed.

@@

\begin{center}
\scalebox{1.0}{
  \includegraphics{diagrams/nat-parity.pdf}
}
\end{center}

>>>>

\paragraph*{Universality.} 
As an example, consider the implementation of the Toffoli gate below,
which establishes that {{langRevT}} is complete for combinational
circuits. This encoding uses the type {{1+1}} as {{bool}} and values
{{left ()}} and {{right()}} as {{true}} and {{false}}.  Boolean
negation not is simply {{swap+}}. The Toffoli gate takes three boolean
inputs: if the first two inputs are {{true}} then the third is
negated.

Even though {{langRevT}} lacks conditional expressions, they are
expressible using the distributivity laws as we demonstrate.  Given
any combinator {{c : b <-> b}} we can construct a combinator called
{{if_c : bool*b <-> bool*b}} in terms of {{c}}, where {{if_c}}
behaves like a one-armed if-expression. If the supplied boolean is
true then the combinator {{c}} is used to transform the value of type
{{b}}. If the boolean is {{false}}, then the value of type {{b}}
remains unchanged. We can write down the combinator for {{if_c}} in
terms of {{c}} as {{distrib (;) ((id (*) c) (+) id) (;) factor}}.

Given {{if_c}}, constructing the Toffoli gate is straightforward. If
we choose {{c}} to be {{not}} we get {{if_{not} }} which is
{{cnot:bool*bool<->bool*bool}} and {{if_{cnot} }} is the required
{{toffoli : bool * (bool * bool) <-> bool * (bool * bool)}}, drawn
below.
\begin{center}
\scalebox{1.2}{
  \includegraphics{diagrams/toffoli.pdf}
}
\end{center}
\noindent Previous work~\cite[Sec.~5]{James:2012:IE:2103656.2103667}
establishes that {{langRevT}} is Turing complete by showing the compilation
of a Turing complete language --- a first-order language with numbers and
iteration --- to {{langRevT}}.

@@

\vspace{-20pt}
\begin{small}
\begin{multicols}{2}  
%subcode{bnf} include main
% Numbers, n, m, p = 0 | n + 1
% Machine states = <n, n, n>

@@

\paragraph*{Stuck States.}
The type systems of most languages are not expressive enough to encode the
precise domain and range of a function. For example, in most typed languages,
division by zero is considered type-correct and the runtime system is
required to deal with such an error. A stuck state of an abstract machine is
just a manifestation of this general problem. The common solutions are:
\begin{itemize}
\item \emph{Use a more expressive type system.}  One could augment
  {{langRevT}} with a richer type system that distinguishes non-zero numbers
  from those that can be zero, thereby eliminating the {{sub1 0}} situation
  entirely. Similarly, in the meta-circular interpreter in
  Sec.~\ref{sec:langrevt-int}, one could use generalized abstract data types
  (GADTs~\cite{phantom,Xi:2003:GRD:604131.604150}) to eliminate the stuck
  states.

@@

We show the generality of our construction by applying it to three
non-trivial examples: a tree traversal, parity translation of numbers
and a meta-circular interpreter for {{langRevT}}.

@@

Deriving {{langRevT}} combinators from abstract machines can sometimes
be used as an efficient indirect way of programming in {{langRevT}}.
It is well known that every number {{n}} can be represented as
{{2a+0}} or {{2a+1}} depending on whether it is odd or even.  The
later can be viewed as the algebraic data type
{{par=odd~|~even~|~A~par}} such that 
{{0 = even, 1 = odd, 2 = A even, 3 = A odd, 4 = A (A even)}} etc where
the nesting of {{A}} constructor indicates the value of {{a}}.

Say we wish to build a {{langRevT}} combinator to map {{nat}} to its
parity encoded version {{par}}, expressible as the {{langRevT}} type
{{@@x.(1+1)+x}}. Since {{par}} is not a fixed unfolding of {{nat}}
({{@@x.1+x}}), it is not immediately apparent how such a combinator
can be constructed.

@@

\begin{center}
\scalebox{1.0}{
  \includegraphics{diagrams/nat-parity2.pdf}
}
\end{center}
